Parsing Postfix log files

### 2008-07-09 doc/logparser-other-parsers-review.tex
Corrections from Mads.

    \item A minor difference is that most parsers don't handle compressed
        files; both \parsername{} and Splunk handle them transparently,
        without user intervention; Sawmill and Lire can be configured to
        support compressed files, but Sawmill exhibits a dramatic slow down
        when thus configured.  Although this is a minor difference, support
        for reading compressed log files is quite helpful, as it
        dramatically reduces the disk space required to store old log
        files.

@@

Isoqlog produces a report listing the number of mails sent by each unique
sender address, and separately the total number of bytes transferred; both
reports are produced for daily, monthly and annual time spans, but only for
the domains listed in its configuration file (making it impossible to
produce reports for every sender domain).  It appears to ignore all log
lines except for those for the current day, though it does maintain a
record of data previously extracted, which the newly extracted data is
merged into (no information is provided on the format of the data store).
The data extracted appears to be limited to the number of mails sent by
each sender, unlike the copious amounts of data extracted by \parsername{}.
It doesn't utilise rejection log lines in any way, so is unsuitable for the
purposes of this project.  Its parsing is completely inextensible, indeed
is almost incomprehensible, relying on \texttt{scanf(3)}, fixed offsets and
low level string manipulation; it is the opposite end of the spectrum to
\parsernames{} parsing.  It doesn't handle any of the complications
discussed in this document, doesn't gather the breadth of data required for
this project, and ignores the majority of log lines produced by Postfix.
It ignores all log lines earlier than today, so testing with the test log
files is pointless.

@@

This utility tracks the number of events which occurred over a particular
time and warns if the frequency of events passes a certain threshold.  It's
designed to provide real time alerts when dictionary attacks, mail loops or
similar problems occur. It doesn't attempt to extract or save data,
correlate log lines, or any of the more advanced tasks described in this
document, because it is not designed to do so.  The user must construct the
\regexes{} to match significant input lines before it will parse any log
files.  This program was not reviewed in further detail because its aims
are so far removed from the aims of this project.

@@

Anteater doesn't have any English documentation so it's difficult, nigh
impossible, for this author to accurately comment on what analysis it
performs.  It did not run successfully when tested, and its parsing would
certainly be out of date as Postfix has evolved considerably since this
tool was last updated (November 2003).  As it neither ran successfully nor
has documentation the author can read a detailed review cannot be provided.

@@

It also counts the number of log lines parsed and skipped.  Log lines are
correlated based on the queueid (referred to as sessionname [sic] within
Logrep), but not by \pid{}.  The parsing is error prone: empty fields are
saved when the log line doesn't match the \regex{}, though it appears that
they will not overwrite existing data.  Most notably rejections are
completely ignored, making it unsuitable for the purposes of this project.
It doesn't attempt to address any of the complications in
\sectionref{complications} \sectionref{additional complications} except for
correlating by queueid.

@@

Logrep extracts far less data from Postfix log files than \parsername{},
completely ignores rejections, is effectively undocumented, doesn't deal
with the more complicated aspects of Postfix log files, and at the time of
writing doesn't work properly.

@@

While there are other programs available which perform basic Postfix log
parsing (some to a greater level of detail than others), few attempt to
correlate log lines by queueid (none correlate by \pid{}) to produce an
overall record of the journey each mail traverses through Postfix.  None of
the reviewed parsers collect the breadth of information gathered by
\parsername{}, nor make it as easy to extend the parser to handle new log
lines.  Most other parsers immediately generate a report and discard the
data extracted from the log files; those which don't discard the data
retain it in a format inaccessible to other tools.  Nearly all of the
parsers reviewed can produce a report of greater or lesser detail and
complexity, a facility not offered by \parsername{}; reporting is deferred
to a subsequent application.  The quality of the documentation offered by
the subset of parsers which provide some varies from unusable to good; none
of the parsers provide any documentation on the format of their data stores
(if any).  Less than half of the parsers were capable of parsing the
\numberOFlogFILES{} test log files, and improving or extending the parsers
would have been quite a difficult task for any of the parsers.  Table
\refwithpage{Summary of parsers' features} provides a summary of the
parsers' features.

>>>>

    \item A minor difference is that most parsers do not handle compressed
        files; both \parsername{} and Splunk handle them transparently,
        without user intervention; Sawmill and Lire can be configured to
        support compressed files, but Sawmill exhibits a dramatic slow down
        when thus configured.  Although this is a minor difference, support
        for reading compressed log files is quite helpful, as it
        dramatically reduces the disk space required to store old log
        files.

@@

Isoqlog produces a report listing the number of mails sent by each unique
sender address, and separately the total number of bytes transferred; both
reports are produced for daily, monthly and annual time spans, but only for
the domains listed in its configuration file (making it impossible to
produce reports for every sender domain).  It appears to ignore all log
lines except for those for the current day, though it does maintain a
record of data previously extracted, which the newly extracted data is
merged into (no information is provided on the format of the data store).
The data extracted appears to be limited to the number of mails sent by
each sender, unlike the copious amounts of data extracted by \parsername{}.
It does not utilise rejection log lines in any way, so is unsuitable for
the purposes of this project.  Its parsing is completely inextensible,
indeed is almost incomprehensible, relying on \texttt{scanf(3)}, fixed
offsets and low level string manipulation; it is the opposite end of the
spectrum to \parsernames{} parsing.  It does not handle any of the
complications discussed in this document, does not gather the breadth of
data required for this project, and ignores the majority of log lines
produced by Postfix.  It ignores all log lines earlier than today, so
testing with the test log files is pointless.

@@

This utility tracks the number of events which occurred over a particular
time and warns if the frequency of events passes a certain threshold.  It
is designed to provide real time alerts when dictionary attacks, mail loops
or similar problems occur. It does not attempt to extract or save data,
correlate log lines, or any of the more advanced tasks described in this
document, because it is not designed to do so.  The user must construct the
\regexes{} to match significant input lines before it will parse any log
files.  This program was not reviewed in further detail because its aims
are so far removed from the aims of this project.

@@

Anteater does not have any English documentation so it is difficult, nigh
impossible, for this author to accurately comment on what analysis it
performs.  It did not run successfully when tested, and its parsing would
certainly be out of date as Postfix has evolved considerably since this
tool was last updated (November 2003).  As it neither ran successfully nor
has documentation the author can read a detailed review cannot be provided.

@@

It also counts the number of log lines parsed and skipped.  Log lines are
correlated based on the queueid (referred to as sessionname [sic] within
Logrep), but not by \pid{}.  The parsing is error prone: empty fields are
saved when the log line does not match the \regex{}, though it appears that
they will not overwrite existing data.  Most notably rejections are
completely ignored, making it unsuitable for the purposes of this project.
It does not attempt to address any of the complications in
\sectionref{complications} \sectionref{additional complications} except for
correlating by queueid.

@@

Logrep extracts far less data from Postfix log files than \parsername{},
completely ignores rejections, is effectively undocumented, does not deal
with the more complicated aspects of Postfix log files, and at the time of
writing does not work properly.

@@

While there are other programs available which perform basic Postfix log
parsing (some to a greater level of detail than others), few attempt to
correlate log lines by queueid (none correlate by \pid{}) to produce an
overall record of the journey each mail traverses through Postfix.  None of
the reviewed parsers collect the breadth of information gathered by
\parsername{}, nor make it as easy to extend the parser to handle new log
lines.  Most other parsers immediately generate a report and discard the
data extracted from the log files; those which do not discard the data
retain it in a format inaccessible to other tools.  Nearly all of the
parsers reviewed can produce a report of greater or lesser detail and
complexity, a facility not offered by \parsername{}; reporting is deferred
to a subsequent application.  The quality of the documentation offered by
the subset of parsers which provide some varies from unusable to good; none
of the parsers provide any documentation on the format of their data stores
(if any).  Less than half of the parsers were capable of parsing the
\numberOFlogFILES{} test log files, and improving or extending the parsers
would have been quite a difficult task for any of the parsers.  Table
\refwithpage{Summary of parsers' features} provides a summary of the
parsers' features.

### 2008-07-09 doc/logparser-parsing-algorithm.tex
Added graph showing distribution of rules per action.
Corrections from Mads.

\section{Parsing algorithm}

@@

Where the rules are quite simple and each rule is completely independent of
its fellows, the algorithm is significantly more complicated and highly
internally interdependent.  The algorithm deals with all the complications
of parsing, the eccentricities and oddities of Postfix log files,
presenting the resulting data in a normalised, easy to use representation.
The algorithm's task is to follow the journey each mail takes through
Postfix, combining the data extracted by rules into a coherent whole,
saving it in a useful and consistent form, and performing housekeeping
duties.

@@

    \item Follow the progress of the accepted mail until it's either
        delivered, bounced or deleted, then log all data, and finish.

@@

Multiple independent mails may be delivered during one connection: this
requires the algorithm to clone the current data as soon as a mail is
accepted, so that subsequent mails won't trample over each other's data.
This must be done every time a mail is accepted, as it's impossible to tell
in advance which connections will accept multiple mails.  Happily, once the
mail has been accepted log entries won't be correlated by \pid{} for that
mail any more (its queueid will be used instead), so there isn't any
ambiguity about which mail a given log line belongs
to.\footnote{Unfortunately this statement is not completely accurate: see
\sectionref{timeouts-during-data-phase} for details.  However in general
there isn't any ambiguity about which data structure should be used for a
given log line.}  The original connection will be discarded unsaved when
the client disconnects if it doesn't have any data worth saving, i.e.\ no
rejections.  One unsolved difficulty is distinguishing between different
groups of rejections, e.g.\ when dealing with the following sequence:

@@

Unfortunately, while all log lines from an individual process appear in
chronological order, the order in which log lines from different processes
are interleaved is subject to the vagaries of process scheduling.  In
addition, the first log line belonging to the child mail (the log line
cited above belongs to the parent mail) is logged by \daemon{qmgr}, so the
order also depends on how busy \daemon{qmgr} is.\footnote{Postfix is quite
paranoid about mail delivery, an excellent characteristic for an \MTA{} to
possess, so it won't log that the child has been created until it is
absolutely certain that the mail has been written to disk.}

@@

    \item The client disconnects (action: DISCONNECT\@; flowchart:~6).  If
        Postfix has rejected any \SMTP{} commands the data will be saved to
        the database; if not there won't be any data to save (any mails
        accepted will already have been cloned so their data is in another
        data structure).

@@

    \item If the mail is of unknown origin it is assumed to be a child mail
        whose parent hasn't yet been identified (action: COMMIT\@;
        flowchart:~15).  Mark the mail as ready for entry in the database
        (flowchart:~16), and wait for the parent to deal with it
        (flowchart:~17).  The mail should not have subsequent log entries;
        only its parent should refer to it.

@@

    \item [IGNORE] This rule just returns successfully; it is used when a
        line needs to be parsed for completeness but doesn't either provide
        any useful data or require anything to be done.

@@

    \item [TIMEOUT] The connection timed out so the mail currently being
        transferred must be discarded. The mail may have been accepted, in
        which case there's a data structure to dispose of, or it may not in
        which case there isn't.  See
        \sectionref{timeouts-during-data-phase} for the gory details.

@@

@@

    \item Is \daemon{smtpd} the only Postfix component that produced a log
        line for this mail?  Every mail which passes normally through
        Postfix will have a \daemon{cleanup} line, and later a
        \daemon{qmgr} log line; lack of a \daemon{cleanup} line is a sure
        sign the mail didn't make it too far.

@@

Some mail clients disconnect abruptly if a second or subsequent recipient
is rejected; they may also disconnect after other errors, but such
disconnections are either unimportant or are handled elsewhere in the
parser (\sectionref{timeouts-during-data-phase}).  Sadly, Postfix doesn't
log a message saying the mail has been discarded, as should be expected by
now.  The checks to identify this happening are:

@@

    \item Is the mail missing its \daemon{cleanup} log message?  Every mail
        which passes through Postfix will have a \daemon{cleanup} line;
        lack of a \daemon{cleanup} line is a sure sign the mail didn't make
        it too far.

@@

    \item The queueid must not have been reused yet, i.e.\ there isn't an
        entry in the state tables for the queueid.

@@

The solution to this problem is to examine the list of programs which have
logged lines for each mail, comparing the list against a table of
known-good combinations of programs (this check is performed during the
COMMIT action).  If the mail's combination is found in the valid list the
mail can be entered in the database; if the combination is not found entry
must be postponed and the mail flagged for later entry.  The
SAVE\_BY\_QUEUEID action checks for that flag and retries entry if it's
found; if the additional log lines have caused the mail to reach a valid
combination entry will proceed, otherwise it must be postponed once more.

@@

Postfix logs the recipient address when delivery of a mail is attempted, so
if delivery has yet to be attempted the parser cannot determine the
recipient address or addresses.  This is a problem when mail is arriving
faster than Postfix can attempt delivery, and the administrator deletes
some of the mail (because it's the result of a mail loop, mail bomb, or
joe~job) before Postfix has had a chance to try to deliver it.  In this
case the recipient address will not have been logged, so a dummy recipient
address needs to be added, as every mail is required to have at least one
recipient.  This complication has been observed in few log files, but
typically when it does occur there will be many instances of it, closely
grouped.  The DELETE action is responsible for handling this complication.

@@

This is yet another complication which only occurs during periods of
extremely high load, when process scheduling vagaries and even hard disk
access times cause strange behaviour.  In this complication bounce
notification mails are created, delivered and deleted from the queue,
\textit{before\/} the log line from \daemon{bounce} is logged.  To deal
with this the COMMIT action maintains a cache of recently committed bounce
notification mails, which the BOUNCE action subsequently checks if the
bounce mail isn't already in the state tables. If a mail exists in the
cache under the correct queueid, and its start time is less than ten
seconds before the timestamp of the bounce log line, it is assumed that the
bounce notification mail has already been processed and the BOUNCE action
does not create one.  If the queueid exists it is removed from the cache,
as it has either just been used or it is too old to use in future.  Whether
the BOUNCE action creates a mail or finds one existing in the state tables,
it flags the mail as having been seen by the BOUNCE action; if this flag is
present the COMMIT action will not add the mail to the cache of recent
bounce notification mails.\footnote{This is not required to correctly deal
with the complication, but is an optimisation to reduce memory usage of the
parser; on the occasions the author has observed this action occurring
there have been a huge number of bounce notification mails generated --- if
every bounce notification mail was cached it would dramatically increase
the memory requirements of the parser.}  The cache of bounce notification
mails will be pruned whenever the parser's state is saved, though if the
size of the cache ever becomes a problem it could be pruned periodically to
keep the size in check.

@@

\subsection{Conclusion}

This section presented the core of the parser, starting with a very high
level view and the initial complications which arose.  A flow chart showing
the paths a mail may take through the nascent simplified algorithm was
provided, followed by an explanation of those paths, and a discussion of
the parser's emergent behaviour --- the data from the log files creates the
paths in the flow chart, they aren't specified anywhere in the parser.  The
framework which holds the parser together was covered next, after which
came a description of the current actions provided by the parser, and the
algorithm for analysing unparsed log lines to create \regexes{} for new
rules.  Detecting, diagnosing and defeating complications forms the largest
single portion of this document, mirroring the development of the parser,
The complications are described in the order they were overcome, with
subsequent problems affecting fewer mails (often by an order or magnitude),
though the time required to solve problems increased with each successive
problem.

>>>>

\section{Parsing Algorithm}

@@

Whereas the rules are quite simple and each rule is completely independent
of the other rules, the algorithm is significantly more complicated and
highly internally interdependent.  The algorithm deals with all the
complications of parsing, the eccentricities and oddities of Postfix log
files, presenting the resulting data in a normalised, easy to use
representation.  The algorithm's task is to follow the journey each mail
takes through Postfix, combining the data extracted by rules into a
coherent whole, saving it in a useful and consistent form, and performing
housekeeping duties.

@@

    \item Follow the progress of the accepted mail until it is either
        delivered, bounced or deleted, then log all data, and finish.

@@

Multiple independent mails may be delivered during one connection: this
requires the algorithm to clone the current data as soon as a mail is
accepted, so that subsequent mails will not trample over each other's data.
This must be done every time a mail is accepted, as it is impossible to
tell in advance which connections will accept multiple mails.  Happily,
once the mail has been accepted log entries will not be correlated by
\pid{} for that mail any more (its queueid will be used instead), so there
is not any ambiguity about which mail a given log line belongs
to.\footnote{Unfortunately this statement is not completely accurate: see
\sectionref{timeouts-during-data-phase} for details.  However in general
there is not any ambiguity about which data structure should be used for a
given log line.}  The original connection will be discarded unsaved when
the client disconnects if it does not have any data worth saving, i.e.\ no
rejections.  One unsolved difficulty is distinguishing between different
groups of rejections, e.g.\ when dealing with the following sequence:

@@

Unfortunately, while all log lines from an individual process appear in
chronological order, the order in which log lines from different processes
are interleaved is subject to the vagaries of process scheduling.  In
addition, the first log line belonging to the child mail (the log line
cited above belongs to the parent mail) is logged by \daemon{qmgr}, so the
order also depends on how busy \daemon{qmgr} is.\footnote{Postfix is quite
paranoid about mail delivery, an excellent characteristic for an \MTA{} to
possess, so it will not log that the child has been created until it is
absolutely certain that the mail has been written to disk.}

@@

    \item The client disconnects (action: DISCONNECT\@; flowchart:~6).  If
        Postfix has rejected any \SMTP{} commands the data will be saved to
        the database; if not there will not be any data to save (any mails
        accepted will already have been cloned so their data is in another
        data structure).

@@

    \item If the mail is of unknown origin it is assumed to be a child mail
        whose parent has not yet been identified (action: COMMIT\@;
        flowchart:~15).  Mark the mail as ready for entry in the database
        (flowchart:~16), and wait for the parent to deal with it
        (flowchart:~17).  The mail should not have subsequent log entries;
        only its parent should refer to it.

@@

    \item [IGNORE] This rule just returns successfully; it is used when a
        line needs to be parsed for completeness but does not either
        provide any useful data or require anything to be done.

@@

    \item [TIMEOUT] The connection timed out so the mail currently being
        transferred must be discarded. The mail may have been accepted, in
        which case there's a data structure to dispose of, or it may not in
        which case there is not.  See
        \sectionref{timeouts-during-data-phase} for the gory details.

@@

The distribution of rules per action is shown in
graph~\refwithpage{Distribution of rules per action}.

@@

    \item Is \daemon{smtpd} the only Postfix component that produced a log
        line for this mail?  Every mail which passes normally through
        Postfix will have a \daemon{cleanup} line, and later a
        \daemon{qmgr} log line; lack of a \daemon{cleanup} line is a sure
        sign the mail did not make it too far.

@@

Some mail clients disconnect abruptly if a second or subsequent recipient
is rejected; they may also disconnect after other errors, but such
disconnections are either unimportant or are handled elsewhere in the
parser (\sectionref{timeouts-during-data-phase}).  Sadly, Postfix does not
log a message saying the mail has been discarded, as should be expected by
now.  The checks to identify this happening are:

@@

    \item Is the mail missing its \daemon{cleanup} log message?  Every mail
        which passes through Postfix will have a \daemon{cleanup} line;
        lack of a \daemon{cleanup} line is a sure sign the mail did not
        make it too far.

@@

    \item The queueid must not have been reused yet, i.e.\ there is not an
        entry in the state tables for the queueid.

@@

The solution to this problem is to examine the list of programs which have
logged lines for each mail, comparing the list against a table of
known-good combinations of programs (this check is performed during the
COMMIT action).  If the mail's combination is found in the valid list the
mail can be entered in the database; if the combination is not found entry
must be postponed and the mail flagged for later entry.  The
SAVE\_BY\_QUEUEID action checks for that flag and retries entry if it is
found; if the additional log lines have caused the mail to reach a valid
combination entry will proceed, otherwise it must be postponed once more.

@@

Postfix logs the recipient address when delivery of a mail is attempted, so
if delivery has yet to be attempted the parser cannot determine the
recipient address or addresses.  This is a problem when mail is arriving
faster than Postfix can attempt delivery, and the administrator deletes
some of the mail (because it is the result of a mail loop, mail bomb, or
joe~job) before Postfix has had a chance to try to deliver it.  In this
case the recipient address will not have been logged, so a dummy recipient
address needs to be added, as every mail is required to have at least one
recipient.  This complication has been observed in few log files, but
typically when it does occur there will be many instances of it, closely
grouped.  The DELETE action is responsible for handling this complication.

@@

This is yet another complication which only occurs during periods of
extremely high load, when process scheduling vagaries and even hard disk
access times cause strange behaviour.  In this complication bounce
notification mails are created, delivered and deleted from the queue,
\textit{before\/} the log line from \daemon{bounce} is logged.  To deal
with this the COMMIT action maintains a cache of recently committed bounce
notification mails, which the BOUNCE action subsequently checks if the
bounce mail is not already in the state tables. If a mail exists in the
cache under the correct queueid, and its start time is less than ten
seconds before the timestamp of the bounce log line, it is assumed that the
bounce notification mail has already been processed and the BOUNCE action
does not create one.  If the queueid exists it is removed from the cache,
as it has either just been used or it is too old to use in future.  Whether
the BOUNCE action creates a mail or finds one existing in the state tables,
it flags the mail as having been seen by the BOUNCE action; if this flag is
present the COMMIT action will not add the mail to the cache of recent
bounce notification mails.\footnote{This is not required to correctly deal
with the complication, but is an optimisation to reduce memory usage of the
parser; on the occasions the author has observed this action occurring
there have been a huge number of bounce notification mails generated --- if
every bounce notification mail was cached it would dramatically increase
the memory requirements of the parser.}  The cache of bounce notification
mails will be pruned whenever the parser's state is saved, though if the
size of the cache ever becomes a problem it could be pruned periodically to
keep the size in check.

@@

\subsection{Summary}

This section presented the core of the parser, starting with a very high
level view and the initial complications which arose.  A flow chart showing
the paths a mail may take through the nascent simplified algorithm was
provided, followed by an explanation of those paths, and a discussion of
the parser's emergent behaviour --- the data from the log files creates the
paths in the flow chart, they are not specified anywhere in the parser.
The framework which holds the parser together was covered next, after which
came a description of the current actions provided by the parser, and the
algorithm for analysing unparsed log lines to create \regexes{} for new
rules.  Detecting, diagnosing and defeating complications forms the largest
single portion of this document, mirroring the development of the parser,
The complications are described in the order they were overcome, with
subsequent problems affecting fewer mails (often by an order or magnitude),
though the time required to solve problems increased with each successive
problem.

### 2008-07-09 doc/logparser-parsing-rules.tex
Corrections from Mads.

\section{Parsing rules}

@@

\subsection{Introduction}

@@

% Don't reformat this!
\begin{tabular}[]{ll}

@@

    \item [program and regex:] If the program in the rule equals the
        program which logged the log line a match using the \regex{} will
        be attempted against the log line; if the match is successful the
        action will be executed, if not the next rule will be tried.  If
        the program-specific rules don't match the log line, the parser
        will fall back to generic rules; if those rules are unsuccessful a
        warning will be issued.

>>>>

\section{Parsing Rules}

@@

@@

% do not reformat this!
\begin{tabular}[]{ll}

@@

    \item [program and regex:] If the program in the rule equals the
        program which logged the log line a match using the \regex{} will
        be attempted against the log line; if the match is successful the
        action will be executed, if not the next rule will be tried.  If
        the program-specific rules do not match the log line, the parser
        will fall back to generic rules; if those rules are unsuccessful a
        warning will be issued.
